A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-11-11 inst/manuscript/manuscript.tex
section 1 and 2 jay updates

  \journal{TBD} % Sets Journal name

@@

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

@@

  \title{A comparison of design-based and model-based approaches for finite
population spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matt Higham}

@@

      \address[USEPA]{United States Environmental Protection Agency, 200 SW 35th St,
Corvallis, Oregon, 97333}
    \address[STLAW]{Saint Lawrence University Department of Mathematics, Computer Science,
and Statistics, 23 Romoda Drive, Canton, New York, 13617}
    \address[OSU]{Oregon State University Department of Statistics, 239 Weniger Hall,
Corvallis, Oregon, 97331}
    \address[NOAA]{Marine Mammal Laboratory, Alaska Fisheries Science Center, National
Oceanic and Atmospheric Administration, Seattle, Washington, 98115}
      \cortext[1]{Corresponding Author}

@@

\emph{Text based on elsarticle sample manuscript, see
\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}

Potential Journals:

\begin{itemize}
\tightlist
\item
  Ecological Applications
\item
  Methods in Ecology and Evolution
\item
  Journal of Applied Ecology
\item
  Environmetrics
\item
  Environmental and Ecological Statistics
\end{itemize}

@@

There are two general approaches for using data to make statistical
inferences about a population: design-based and model-based. When data
cannot be obtained for all units in a population (population units),
data on a subset of the population units is collected and called a
sample. In the design-based approach, inferences about the underlying
population are informed from a probabilistic process in which population
units are selected to be in the sample. Alternatively, in the
model-based approach, inferences are made from specific assumptions
about the underlying process that generated the data. Each paradigm has
a deep historical context (Sterba, 2009) and its own set of general
advantages (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Thereafter, several comparisons between
design-based and model-based for spatial data have been considered (Brus
and De Gruijter, 1997; Ver Hoef, 2002, 2008). Cooper (2006) review the
two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design and
model-based principles (see e.g.~Cicchitelli and Montanari (2012),
Chan-Golston et al. (2020) for a Bayesian approach, and Sterba (2009)).
More recent overviews include Brus (2020) and Wang et al. (2012).

Though comparisons between design-based and model-based approaches to
spatial data have been studied, no numerical comparison has been made
between design-based approaches that incorporate spatial locations and
model-based approaches. In this manuscript, we compare design-based
approaches that incorporate spatial locations to model-based approaches
for spatial data. Though these comparisons generalize to both finite
populations (e.g.~point resources) and infinite populations (e.g.~linear
and areal resources), we focus on applications to finite populations.
The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we compare sampling and estimation procedures
between the design-based approach and the model-based approach for
spatial data. In Section \ref{sec:numstudy}, we use a simulation
approach to study the behavior and performance of both approaches. In
Section \ref{application}, we use both approaches to analyze real data.
And in Section \ref{sec:discussion}, we end with a discussion and
provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame according to
a sampling design. A sampling frame is the set of all units available to
be sampled. A sampling design assigns a positive probability of
inclusion (inclusion probability) to each unit in the sampling frame.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sample, and cluster sampling. These sampling
designs tend to select units from the sampling frame independently of
other units, so we call them ``Independent Random Sampling'' (IRS)
designs. Sampling designs incorporating the spatial locations of units
in the sample frame are called spatially balanced designs. Spatially
balanced designs can be obtained using the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004),
which we discuss in more detail in Section \ref{subsec:spb_design}. The
design-based approach combines the randomness of the sampling design and
the data collected via the sample to estimate parameters (e.g.~means and
totals) of a population. Generally, these population parameters are
assumed to be fixed, unknown constants.

@@

The model-based approach assumes the data are a random realization of a
data-generating process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement random sampling.

Instead of estimating fixed but unknown parameters (as in the
design-based approach), the goal of model-based inference in the spatial
context is often to predict a realized variable, or value. For example,
suppose the realized mean of all population units is the value of
interest. Instead of \emph{estimating} a fixed, unknown mean, we are
\emph{predicting} the value of the mean, a random variable. Prediction
intervals are then derived leveraging assumptions of the data generating
process. If we repeatedly generate the response values from a fixed
spatial process and obtained a sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data. A visual comparison of the design-based and
model-based assumptions is provided in Figure \ref{fig:fig1} (Brus
(2020) provides a similar figure).

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples of n = 4. The response values at each site are fixed, but we obtain different estimates for the mean response because the randomly sampled sites vary from sample to sample. In the bottom row, we have three realizations of the same spatial process sampled at the same locations. The spatial process generating the response values has a single mean, but the realized mean is different in each of the three panels.}\label{fig:fig1}
\end{figure}

@@

Spatially balanced samples can be obtained using the design-based
approach. Spatially balanced samples are useful because parameter
estimates from these samples tend to vary less than parameter estimates
from samples that are not spatially balanced (Barabesi and Franceschi,
2011; Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson
et al., 2013; Stevens and Olsen, 2004; Wang et al., 2013). The first
spatially balanced sampling algorithm that saw widespread use was the
Generalized Random Tessellation Stratified (GRTS) algorithm (Stevens and
Olsen, 2004). To quantify the spatial balance of a sample, Stevens and
Olsen (2004) proposed loss metrics based on Voroni polygons. Since GRTS
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use Generalized Random Tessellation Stratified
(GRTS) sampling because it has several attractive properties: GRTS
sampling accommodates finite and infinite sampling frames; accommodates
equal, unequal, and proportional (to) size inclusion probabilities;
accommodates legacy (historical) sampling; accommodates a minimum
distance between units in a sample; accommodates reverse hierarchically
ordered replacement units in a sample (replacement units are units
available to be sampled if an original unit cannot be sampled); and is
available in the spsurvey R package Dumelle et al. (2021).

The GRTS algorithm samples from finite and infinite populations by
utilizing a mapping between two-dimensional and one-dimensional space.
The units in the two-dimensional sampling frame are divided into cells
using a hierarchical address. This hierarchical address is then used to
map the units from two-dimensional space to a one-dimensional line where
each unit's line length equals its inclusion probability. A systematic
sample is conducted on the line and linked back to a unit in
two-dimensional space, which results in the desired sample. Stevens and
Olsen (2004) and Dumelle et al. (2021) provide further details.

After selecting a GRTS sample, data are collected and used to estimate
population parameters. To unbiasedly estimate population means and
totals from sample data, one can use the Horvitz-Thompson estimator
(Horvitz and Thompson, 1952). If \(\tau\) is a population total, the
Horvitz-Thompson estimate of \(\tau\), denoted by \(\hat{\tau}_{ht}\),
is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th unit in the sample
and \(\pi_i\) is the inclusion probability of the \(i\)th unit in the
sample. An estimate of the population mean is obtained by dividing
\(\hat{\tau}_{ht}\) by the population size.

While the Horvitz-Thompson estimator is unbiased for population means
and totals, it is also important to quantify the uncertainty in these
estimates. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but they have two drawbacks. First,
these estimators rely on calculating \(\pi_{ij}\), the probability that
unit \(i\) and unit \(j\) are both in the sample -- this quantity can be
challenging if not impossible to calculate analytically. Second, these
estimators ignore the spatial locations of the units in the sampling
frame. To address these two drawbacks simultaneously, Stevens and Olsen
(2003) proposed the local neighborhood variance estimator. The local
neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends reduce \(\text{Var}(\hat{\tau})\) compared to
variance estimators ignoring spatial locations, yielding narrower
confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of basing inference off of a specific
sampling design, we assume the data are generated by a spatial process.
Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be a response vector at locations \(s_1\), \(s_2\), . . . , \(s_N\) that
can be measured at the \(N\) population units and is represented as an
\(N \times 1\) vector. Suppose we want to predict some linear function
of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights. For
example, if we want to predict the population total across all
population units, then we would use a vector of 1's for the weights.

We often only have a sample of the \(N\) population units. Denoting
quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with a subscript \emph{u},

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation} where \(\mathbf{X}_s\) and \(\mathbf{X}_u\) are the
design matrices for the sampled and unsampled population units,
respectively; \(\beta\) is the parameter vector of fixed effects; and
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for the
sampled and unsampled population units, respectively. Denoting
\(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), we assume
the expectation of \(\bm{\delta}\) equals \(\mathbf{0}\).

In addition to assuming the expectation of \(\bm{\delta}\) equals
\(\mathbf{0}\), we also assume that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th entry for
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{ps}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{ps} + \sigma^2_n & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{ps}\) is the partial sill measuring
coarse-scale (correlated) variability, \(\sigma^2_{n}\) is the nugget
measuring fine-scale (independent) variability, \(\phi\) is the range
parameter measuring the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). Any spatial covariance function could be used in the place of the
exponential, however, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in (Ver Hoef, 2008), we
note here that the predictor and its variance are both moment-based,
meaning they don't rely on any distributional assumptions.

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. However,
k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013),
random forest (Breiman, 2001), Bayesian models (Chan-Golston et al.,
2020), among others, can also be used to obtain predictions for a mean
or total from spatially correlated responses of a finite population. We
choose to use FPBK because it is faster than a Bayesian approach and
random forest and because Ver Hoef and Temesgen (2013) showed that the
method outperforms k-nearest-neighbors in many scenarios.

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T.-l., 2010. Geostatistical inference
under preferential sampling. Journal of the Royal Statistical Society:
Series C (Applied Statistics) 59, 191--232.

\leavevmode\hypertarget{ref-dumelle2021spsurvey}{}%
Dumelle, M., Olsen, A.R., Kincaid, T., Weber, M., 2021. Selecting and
analyzing spatial probability samples in r using spsurvey. Manuscript
Submitted for Publication.

@@

>>>>

  \journal{Methods in Ecology and Evolution} % Sets Journal name

@@

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

@@

  \title{A comparison of design-based and model-based approaches for
finite population spatial data.}
    \author[USEPA]{Michael Dumelle\corref{1}}

    \author[STLAW]{Matt Higham}

@@

      \address[USEPA]{United States Environmental Protection Agency, 200
SW 35th St, Corvallis, Oregon, 97333}
    \address[STLAW]{Saint Lawrence University Department of Mathematics,
Computer Science, and Statistics, 23 Romoda Drive, Canton, New York,
13617}
    \address[NOAA]{Marine Mammal Laboratory, Alaska Fisheries Science
Center, National Oceanic and Atmospheric Administration, Seattle,
Washington, 98115}
    \address[OSU]{Oregon State University Department of Statistics, 239
Weniger Hall, Corvallis, Oregon, 97331}
      \cortext[1]{Corresponding Author: Michael Dumelle
(Dumelle.Michael@epa.gov)}

@@

@@

There are two general approaches for using data to make frequentist
statistical inferences about a population: design-based and model-based.
When data cannot be collected for all units in a population (population
units), data are collected on a subset of the population units. This
subset is called a sample. In the design-based approach, inferences
about the underlying population are informed via a probabilistic process
assigning some population units to the sample. Alternatively, in the
model-based approach, inferences are made from specific assumptions
about the underlying process generating the data. Each paradigm has a
deep historical context (Sterba, 2009) and its own set of benefits and
drawbacks (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2020; Ver Hoef, 2002; Ver Hoef,
2008; Wang et al., 2012). Cooper (2006) reviews the two approaches in an
ecological context before introducing a ``model-assisted'' variance
estimator that combines aspects from each approach. In addition to
Cooper (2006), there has been substantial research and development into
estimators that use both design and model-based principles (see e.g.,
Sterba (2009), Cicchitelli and Montanari (2012), Chan-Golston et al.
(2020) for a Bayesian approach).

Though comparisons between design-based and model-based approaches to
spatial data have been studied, no numerical comparison has been made
between design-based approaches that incorporate spatial locations and
model-based approaches. In this manuscript, we compare design-based
approaches that incorporate spatial locations to model-based approaches
for spatial data. We focus on finite populations, but these comparisons
generalize to infinite populations as well. A finite population contains
a finite number of population units; an example is lakes (treated as a
whole with the lake centroid representing location) in the contiguous
United States. An infinite population contains an infinite number of
population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and compare several sampling and
estimation procedures of the design-based and model-based approaches for
finite population spatial data. In Section \ref{sec:numstudy}, we use a
simulation approach to study the behavior and performance of both
approaches. In Section \ref{application}, we use both approaches to
analyze real data consisting of mercury concentration from lakes in the
contiguous United States. And in Section \ref{sec:discussion}, we end
with a discussion and provide directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a smapling frame according to
a sampling design. A sampling frame is the set of all units available to
be sampled. A sampling design assigns a positive probability of
inclusion (inclusion probability) to each unit in the sampling frame.
Some examples of commonly used sampling designs include simple random
sampling, stratified random sampling, and cluster sampling. If a
sampling design selects units from the sampling frame while ignoring
their spatial locations, we call them ``Independent Random Sampling''
(IRS) designs. If a sampling design selects units from the sampling
frame while incorporating their spatial locations, we call them
spatially balanced designs. Spatially balanced designs can be obtained
using the Generalized Random Tessellation Stratified (GRTS) algorithm
(Stevens and Olsen, 2004), which we discuss in more detail in Section
\ref{subsec:spb_design}. The design-based approach combines the
randomness of the sampling design and the data collected via the sample
to estimate fixed, unknown parameters (e.g., means and totals) of a
population.

@@

The model-based approach assumes the data are a random realization of a
data-generating stochastic process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement random sampling.

Instead of estimating fixed but unknown parameters like a mean or total
(as in the design-based approach), the goal of model-based inference in
the spatial context is often to predict a realized variable, or value.
For example, suppose the realized mean of all population units is the
value of interest. Instead of \emph{estimating} a fixed, unknown mean,
we are \emph{predicting} the value of the mean, a random variable.
Prediction intervals are then derived using assumptions of the data
generating process. If we repeatedly generate the response values from
the same spatial process and sample, then 95\% of all 95\% prediction
intervals constructed from a procedure with appropriate coverage will
contain their respective realized means. Cressie (1993) and
Schabenberger and Gotway (2017) provide reviews of model-based
approaches for spatial data. A visual comparison of the design-based and
model-based assumptions is provided in Figure \ref{fig:fig1} (Ver Hoef
(2002) and Brus (2020) provide similar figures).

\begin{figure}
\includegraphics[width=1\linewidth]{manuscript_files/figure-latex/fig1-1} \caption{A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples of size four. The response values at each site are fixed, but we obtain different estimates for the mean response because the randomly sampled sites vary from sample to sample. In the bottom row, we have three realizations of the same spatial process sampled at the same locations. The spatial process generating the response values has a single mean, but the realized mean is different in each of the three panels.}\label{fig:fig1}
\end{figure}

@@

The design-based approach can be used to select samples that are
``well-spread'' in space, or spatially balanced. Spatially balanced
samples are useful because parameter estimates from these samples tend
to vary less than parameter estimates from samples that are not
spatially balanced (Barabesi and Franceschi, 2011; Benedetti et al.,
2017; Grafström and Lundström, 2013; Robertson et al., 2013; Stevens and
Olsen, 2004; Wang et al., 2013). The first spatially balanced sampling
algorithm that saw widespread use was the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004). To
quantify the spatial balance of a sample, Stevens and Olsen (2004)
proposed loss metrics based on Voroni polygons. After the GRTS algorithm
was developed, several other spatially balanced sampling algorithms have
emerged, including the Local Pivotal Method (Grafström et al., 2012;
Grafström and Matei, 2018), Spatially Correlated Poisson Sampling
(Grafström, 2012), Balanced Acceptance Sampling (Robertson et al.,
2013), Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017),
and Halton Iterative Partitioning Sampling (Robertson et al., 2018). In
this manuscript, we use the Generalized Random Tessellation Stratified
(GRTS) algorithm to select spatially balanced samples sampling because
the algorithm has several attractive properties. It accommodates finite
and infinite sampling frames. It accommodates equal, unequal, and
proportional (to size) inclusion probabilities. It accommodates legacy
(historical) sampling (Foster et al., 2017). It accommodates a minimum
distance between units in a sample. It accommodates replacement units in
a sample, which are units that can be sampled in place of an original
unit that can no longer be sampled. Lastly, it can be implemented using
the spsurvey R package Dumelle et al. (2021).

The GRTS algorithm samples from finite and infinite populations by
utilizing a mapping between two-dimensional and one-dimensional space.
The units in the two-dimensional sampling frame are divided into cells
using a hierarchical address. This hierarchical address is then used to
map the units from two-dimensional space to a one-dimensional line where
each unit's line length equals its inclusion probability. A systematic
sample is conducted on the line and linked back to a unit in
two-dimensional space, which results in the desired sample. Stevens and
Olsen (2004) provides further details.

After selecting a spatially balanced sample using the GRTS algorithm
(i.e., a GRTS sample), data are collected and used to estimate
population parameters. To unbiasedly estimate population means and
totals from sample data, one can use the Horvitz-Thompson estimator
(Horvitz and Thompson, 1952). If \(\tau\) is a population total, the
Horvitz-Thompson estimate of \(\tau\), denoted by \(\hat{\tau}_{ht}\),
is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th unit in the sample
and \(\pi_i\) is the inclusion probability of the \(i\)th unit in the
sample. An estimate of the population mean can be obtained by dividing
\(\hat{\tau}_{ht}\) by number of population units.

While the Horvitz-Thompson estimator is unbiased for population means
and totals, it is also important to quantify the uncertainty in these
estimates. Horvitz and Thompson (1952) and Sen (1953) provide variance
estimators for \(\hat{\tau}_{ht}\), but these estimators have two
drawbacks. First, they rely on calculating \(\pi_{ij}\), the probability
that unit \(i\) and unit \(j\) are both in the sample -- this quantity
can be challenging if not impossible to calculate analytically. Second,
these estimators ignore the spatial locations of the units in the
sampling frame. To address these two drawbacks simultaneously, Stevens
and Olsen (2003) proposed the local neighborhood variance estimator. The
local neighborhood variance estimator does not rely on \(\pi_{ij}\) and
incorporates spatial locations -- for technical details see Stevens and
Olsen (2003). Stevens and Olsen (2003) show the local neighborhood
variance estimator tends to reduce the estimated variance of
\(\hat{\tau}\) compared to variance estimators ignoring spatial
locations, yielding narrower confidence intervals for \(\tau\).

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
process. Ver Hoef (2008) gives details on the theory of FPBK, but some
of the basic principles are summarized below. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to predict some linear function of the response variable,
\(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights. For
example, if we want to predict the population total across all
population units, then we would use a vector of 1's for the weights.

We often only have a sample of the \(N\) population units. Denoting
quantities that are part of the sampled population units with a
subscript \emph{s} and quantities that are part of the unsampled
population units with subscript \emph{u}, let

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation} where \(\mathbf{X}_s\) and \(\mathbf{X}_u\) are the
design matrices for the sampled and unsampled population units,
respectively, and \(\bm{\beta}\) is the parameter vector of fixed
effects.

Let \(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), where
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for teh
sampled and unsampled population units, respectively. We assume
\(E(\bm{\delta} = \mathbf{0}\) and that there is spatial correlation in
\(\bm{\delta}\) that can be modeled using a covariance function. It is
common to assume the covariance function is second-order stationary and
isotropic (Cressie, 1993), and that the spatial covariance decreases as
the separation between population units increases. Many spatial
covariance functions exist, but the primary function we use throughout
the simulations and applications in this manuscript is the exponential
covariance function: the \(i,j\)th elment of hte matrix
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is dependent random error variance
measuring coarse-scale (correlated) variability, \(\sigma^2_{n}\) is the
independent random error variance measuring fine-scale (independent)
variability, \(\phi\) is the range parameter measuring the
distance-decay rate of the correlation, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). Often
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are called the partial sill and
nugget, respectively. Any spatial covariance function could be used in
the place of the exponential, including functions that allow for
non-stationarity or anisotropy (Chiles and Delfiner, 1999, pp. 80--93).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(f(\mathbf{b}'\mathbf{z})\) and its prediction variance can
be computed. While details of the derivation are in (Ver Hoef, 2008), we
note here that the predictor and its variance are both moment-based,
meaning that they do not rely on any distributional assumptions.

We note that we only use FPBK in this paper in order to focus more on
comparing the design-based and model-based approaches. Other methods,
such as k-nearest-neighbors (Fix and Hodges, 1951; Ver Hoef and
Temesgen, 2013), random forest (Breiman, 2001), Bayesian models
(Chan-Golston et al., 2020), among others, could also be used to obtain
predictions for a mean or total from spatially correlated responses of a
finite population. We choose to use FPBK because it is faster than a
Bayesian approach and it was developed with theoretically-based variance
estimators of means and totals for spatial data, whereas random forests
and k-nearest-neighbors use ad-hoc variance estimators in most cases
(Ver Hoef and Temesgen, 2013); additionally, FBPK outperformed the other
methods in most scenarios.

@@

\leavevmode\hypertarget{ref-diggle2010geostatistical}{}%
Diggle, P.J., Menezes, R., Su, T., 2010. Geostatistical inference under
preferential sampling. Journal of the Royal Statistical Society: Series
C (Applied Statistics) 59, 191--232.

\leavevmode\hypertarget{ref-dumelle2021spsurvey}{}%
Dumelle, M., Kincaid, T.M., Olsen, A.R., Weber, M.H., 2021. Spsurvey:
Spatial sampling design and analysis.

@@

\leavevmode\hypertarget{ref-foster2017spatially}{}%
Foster, S.D., Hosack, G.R., Lawrence, E., Przeslawski, R., Hedge, P.,
Caley, M.J., Barrett, N.S., Williams, A., Li, J., Lynch, T., others,
2017. Spatially balanced designs that incorporate legacy sites. Methods
in Ecology and Evolution 8, 1433--1442.
